---
title: "Functions Primer"
description: |
  An intro to functions in R
author:
  - name: Eric Ekholm
    affiliation: CCPS Office of Research & Evaluation
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
---

# Overview

Functions! This document provides an overview of functions and how you can use them to consolidate your code.

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center")

#loading useful packages
library(tidyverse)
library(palmerpenguins)
library(janitor)

```

# What is a Function?

A function is a chunk of code that takes one (or more) inputs and returns one output. Pretty much everything you do in R is a function, including all of the following:

- things from the tidyverse, like `mutate()`
- things in base R, like `c()`,
- arithmetic operators, like `+`

You can write your own function using the `function()` function. For example:

```{r}
times_two <- function(x) {
  x*2
}
```

This creates a new function named "times_two" that takes a single argument, x. Inside the body of the function (between the curly braces), we define what this function does, which is multiply x by two.

We can then call this function:

```{r}
times_two(2)
```

```{r}
times_two(x = 100)
```

We can also call this function inside other functions. For example, we can use it inside `mutate()`:

```{r}
penguins %>%
  mutate(body_mass2 = times_two(body_mass_g)) %>%
  select(body_mass_g, body_mass2) %>%
  head()
```


# Extending Our Function 

Let's say we wanted to extend our `times_two()` function. Instead of multiplying x by two, what if we want to multiply x by y?

```{r}
multiply <- function(x, y) {
  x * y
}
```

We now have a function (named `multiply`) with two inputs, x and y. We can call this like we did our previous function:

```{r}
multiply(2, 2)
```

```{r}
multiply(2, 10)
```

We can also pass a vector into x and multiply each element of that vector by y:

```{r}
multiply(1:5, 2) 
```

And we can use this inside another function:

```{r}
penguins %>%
  mutate(bm_fliplen = multiply(body_mass_g, flipper_length_mm)) %>%
  select(body_mass_g, flipper_length_mm, bm_fliplen) %>%
  head()
```


## Setting Default Values

Suppose that, in our multiply function, we want to multiply x by 2 *most of the time* but want people to have the option to provide another value? We can set 2 as the default value for y that people can overwrite if they want.

```{r}
multiply <- function(x, y = 2) {
  x * y
}
```

Now, if we don't provide a y value, our function will just assume we want to multiply by 2

```{r}
multiply(x = 3)
```

But we *can* provide a y value to overwrite this default:

```{r}
multiply(x = 3, y = 3)
```

# Using Functions as Wrappers

One nice thing about functions is that we can "wrap" them around larger chunks of code to either simplify that code or represent it more succinctly. For example, imagine it's common practice for us to take the penguins data, filter down to a specific species of penguin, and then get the average body mass for that penguin. Let's do this with the Adelie penguins first. We could write this code:

```{r}
penguins %>%
  filter(species == "Adelie") %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
```

But suppose we want to do this with each species. Or suppose this is common enough that we want to do it in multiple documents. We might want to wrap this whole chunk of code in a function.

```{r}
average_penguin_mass <- function(filt_species) {
  penguins %>%
    filter(species == filt_species) %>%
    summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
}

average_penguin_mass(filt_species = "Adelie")
```

In the above, we create a new function `average_penguin_mass()` that has one argument: species. We can tell the function which species we want to calculate the avg mass for, and it will do that. Essentially, we're converting our 3 lines of code into 1.

And we can pass in other species names easily:

```{r}
average_penguin_mass("Gentoo")
```


This could be useful for us if we wanted to, for instance, calculate absences by school -- we could write a function that took a "school" argument to filter results down to a specific school.

# Writing Tidyverse Functions

Writing functions that play nicely with the tidyverse can be a little more difficult than the example above indicates. One nice thing about the tidyverse is that you can pass in unquoted variable names to functions, and tidyverse functions know how to handle them. For example, we can pass `body_mass_g` as an unquoted variable to summarize, and it knows what we want:

```{r}
penguins %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
```


When writing a function, you might want to get the mean of whatever variable you pass into that function. However, for some under-the-hood technical reasons, the following won't work:

```{r, eval = FALSE}
penguin_summarize <- function(var) {
  penguins %>%
    summarize(avg = mean(var, na.rm = TRUE))
}

penguin_summarize(body_mass_g)
```


You can read some more about this phenomenon in [this article](https://dplyr.tidyverse.org/articles/programming.html) or in the [metaprogramming section of the Advanced R book](https://adv-r.hadley.nz/metaprogramming.html) if you want, but quick fix is that if you want to write functions that use unquoted variables in the tidyverse, you need to *embrace* these variables by using {{ }}:

```{r}
penguin_summarize <- function(var) {
  penguins %>%
    summarize(avg = mean({{ var }}, na.rm = TRUE))
}

penguin_summarize(body_mass_g)
```

```{r}
penguin_summarize(flipper_length_mm)
```


# Walking through a {ccpsr} Function

To illustrate a function we might actually use, let's walk through the `fetch_grades()` function from `{ccpsr}`. This function lets users fetch the grades data for students for a given year. It also lets users decide if they want grades for all students or only secondary students. Here's the full code for the function (no worries if some of it doesn't make sense):

```{}
fetch_grades <- function(year, secondary_only = TRUE) {

  #check that year is numeric
  if (!is.numeric(year)) {
    rlang::abort(paste0("`year` must be an integer, not ", typeof(year)))
  }

  #check school year value
  if (year < 2014) {
    rlang::abort("Grades are only available from 2014 onward")
  }
  
  #check that secondary_only is logical
  if (!is.logical(secondary_only)) {
    rlang::abort(paste0("`secondary_only` must be logical, not ", typeof(secondary_only)))
  }

  #configure connection to ccps databases
  con <- ccpsr::set_con()
  
  #create our query
  q <- paste0("SELECT gr.*,
           cr.Academic_Type
           FROM [CCPS_Shared].[CCPS_Shared].[RE_Student_GP_Grades] gr 
           LEFT JOIN (
           SELECT COURSE_ID,
           Academic_Type
           FROM [CCPS_Shared].[CCPS_Shared].[RE_Courses]) cr 
           ON gr.Course_Id = cr.COURSE_ID
           WHERE SchYr = ", year)

  #query database and save data
  tmp <- odbc::dbGetQuery(con, q)

  #filter data to only secondary schools if secondary_only option is set to TRUE
  ret <- if (secondary_only == TRUE) {

      dplyr::filter(tmp, SchoolCode %in% ccpsr::secondary_school_codes)

  } else tmp

  ret <- janitor::clean_names(ret)

  #return the 'ret' object
  return(ret)

}

```

## Function Arguments

We can execute the function by calling it like this:

```{r, eval = FALSE}
#library(ccpsr) -- we will need to load the library first
fetch_grades(year = 2018, secondary_only = TRUE)
```

This will fetch students' grades for the 2018 school year, and it will only fetch grades for secondary students. Let's look back at the function definition to understand these arguments.

```{r}
fetch_grades <- function(year, secondary_only = TRUE) {
  #some code in here
}
```

This tells us that we're creating a function named `fetch_grades`, and that this function will have 2 arguments: `year` and `secondary_only`. We can also see that `year` is the first argument and `secondary_only` is the second argument. These arguments get "passed in" to the function and tell it how to perform its calculations, which we'll see shortly. `R` also lets you provide arguments to functions using the position, so we could call the function like this if we wanted:

```{r, eval = FALSE}
fetch_grades(2018, TRUE)
```

Furthermore, in our function definition, we've provided a *default value* of `TRUE` to `secondary_only`. This tells our function that, if we don't provide a value for the `secondary_only` argument, it will use `TRUE`. So the following function call is equivalent to the previous two:

```{r, eval = FALSE}
fetch_grades(2018)
```

## Value Checks

Now that we understand the function arguments, we can dive into the *body* of the function -- everything between the curly braces. This is the code that actually gets executed when you call the function.

Let's take a look at the first few lines of code inside `fetch_grades()`:

```{r, eval = FALSE}
fetch_grades <- function(year, secondary_only = TRUE) {

  #check that year is numeric
  if (!is.numeric(year)) {
    rlang::abort(paste0("`year` must be an integer, not ", typeof(year)))
  }

  #check school year value
  if (year < 2014) {
    rlang::abort("Grades are only available from 2014 onward")
  }
  
  #check that secondary_only is logical
  if (!is.logical(secondary_only)) {
    rlang::abort(paste0("`secondary_only` must be logical, not ", typeof(secondary_only)))
  }
  
  #more code below...
}
```

These three chunks of code make sure that we're passing valid values into our function. The first one checks to make sure that the value of the `year` argument is numeric; the second one checks that the value of the `year` argument is not less than 2014 (since grade data is only available from 2014 onward); and the third one checks to make sure that the value of `secondary_only` is logical (either `TRUE` or `FALSE`). If any of these checks fail, the function will return the associated error message to the user.

It's super important that you write these sorts of checks when you first start writing functions, especially if you're the primary person using your functions. Once other people start using functions you write, it can be helpful to provide some checks and error messages for them.

What is worth understanding here, though, is that these checks are using the values passed into the arguments by the users. For example, note that the check below makes sure that the `year` value is numeric:

```{r, eval = FALSE}
fetch_grades <- function(year, secondary_only = TRUE) {

  #check that year is numeric
  if (!is.numeric(year)) {
    rlang::abort(paste0("`year` must be an integer, not ", typeof(year)))
  }
  
  #more code below...
}
```


Any time we can to use the value of the `year` argument provided by the user in the function, we provide the variable `year` to our code.

## Querying the Database

After the function checks to make sure the arguments are valid, it then connects to and queries the database to return students' grade data via the following code:

```{}
fetch_grades <- function(year, secondary_only = TRUE) {
  #argument checks here...
  
  #configure connection to ccps databases
  con <- ccpsr::set_con()

  #create our query
  q <- paste0("SELECT gr.*,
           cr.Academic_Type
           FROM [CCPS_Shared].[CCPS_Shared].[RE_Student_GP_Grades] gr 
           LEFT JOIN (
           SELECT COURSE_ID,
           Academic_Type
           FROM [CCPS_Shared].[CCPS_Shared].[RE_Courses]) cr 
           ON gr.Course_Id = cr.COURSE_ID
           WHERE SchYr = ", year)

  #query database and save data
  tmp <- odbc::dbGetQuery(con, q)
  
  #more code here...
}
```


The first thing this does is run the `set_con()` function from `{ccpsr}` to connect to our databases. This uses our computers' login credentials to connect "automagically".

Next, we query the database like we typically would, using `dbGetQuery()` from the `{odbc}` package. The one wrinkle here is that instead of writing out a "typical" SQL query, we want to include our `year` variable. We can do this by passing the whole query into the `paste0()` function, which essentially "pastes" strings together. So the last line of the query above (beginning with "WHERE...") will set SchYr (in our SQL query) equal to whatever value of `year` we passed into the function.

If we want, we can show this behavior with the following function:

```{r}
paste_year <- function(year) {
  paste0("Value of year is ", year)
}

paste_year(2018)
```

So anyway, `fetch_grades()` then executes the query and saves the resulting data to a variable named `tmp` (short for "temporary")

## Return Data

So far, we've passed our two arguments (`year` and `secondary_only`) into our function, checked that the argument values are valid, and executed a SQL query that uses the `year` value to retrieve grades only for that year. The last few lines of our function will (optionally) filter the data a bit more and then return the data to the user:

```{r, eval = FALSE}
fetch_grades <- function(year, secondary_only = TRUE) {
  #more code up here...
  
  #filter data to only secondary schools if secondary_only option is set to TRUE
  ret <- if (secondary_only == TRUE) {

      dplyr::filter(tmp, SchoolCode %in% ccpsr::secondary_school_codes)

  } else tmp

  ret <- janitor::clean_names(ret)

  #return the 'ret' object to the user
  return(ret)
}
```


The first thing we're doing in this block of code is looking at the value of the `secondary_only` argument (which, remember, can only be `TRUE` or `FALSE` thanks to our code checks). If the value of `secondary_only` is `TRUE`, we'll take our `tmp` variable (which is the data returned from the query) and further filter it to only contain data for students in secondary schools, then assign this result to a variable called `ret` (short for "return"). To do this, the function uses the `secondary_school_codes` data built into `{ccpsr}`. If the value of `secondary_only` is `FALSE`, the function won't do this filtering and will just assign all of the data in `tmp` to `ret`.

Next, the function will use the `clean_names()` function from the `{janitor}` package on our `ret` variable to make the column names easier to work with.

Finally, we call the `return()` function to return the `ret` variable to the user. Technically, we can just type `ret` in this last line (we don't actually need `return()`), but I think it illustrates that we're returning this value. It's worth noting that `R` functions can only return a single object, but this object can be of any type (e.g. a vector, data frame, list, etc).

et voila, that's our function!

# Further Reading

If you want to get into the weeds of functions, you can read [this chapter in *Advanced R*](https://adv-r.hadley.nz/functions.html).

If you want to learn more about programming (writing functions) with the tidyverse, you can read [this article](https://dplyr.tidyverse.org/articles/programming.html)
