---
title: "Working with Strings"
description: |
  Working with Strings in R
author:
  - name: Eric Ekholm
    affiliation: CCPS Office of Research & Evaluation
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
---

Some snippets to help working with strings.

```{r}
# load packages
library(tidyverse)
library(ccpsr)

schs <- ccpsr::title_i$organization_name
```

The vast majority of the functions we're using will come from `{stringr}`

## Detect a pattern in a string

This will return a vector with values of either TRUE or FALSE. Useful for filtering.

```{r}
x <- str_detect(schs, "Midlo")

sum(x) # will return the total # of trues
```

## Subset a string

`str_subset()` will take a subset of a character vector that matches a pattern

```{r}
str_subset(schs, "Midlo")
```

## Count the number of characters in a string

```{r}
str_length(schs)
```

## Basic Regex

Regex -- or regular expressions -- provide a standardized way to work with text across multiple programming languages. SQL also uses regex, so some of this might be familiar. Here's a [cheat sheet](https://www.rexegg.com/regex-quickstart.html) that I reference, but some very basic ideas:

- `^` indicates the start of a string
- `$` indicates the end of a string

So we can get any schools that start with M via:

```{r}
str_subset(schs, "^M")

```

Or any schools that end with "HS" via:

```{r}
str_subset(schs, "HS$")
```

`.` is a wildcard operator (i.e. it represents anything), and `*` means "0 or more times". You can combine them so that `.*` means "anything"

```{r}
schs[1:5] == str_subset(schs[1:5], ".*")
```

We'll see how this can be useful later.

`\\d` represents a digit. `+` represents "one or more times". We can combine these to, for instance, get strings that have a digit in them:

```{r}
str_subset(schs, "\\d+")
```

To get schools that don't have a digit, we can use the `negate` argument of `str_subset`:

```{r}
str_subset(schs, "\\d+", negate = TRUE)
```

There are a few of these that are useful for us:

- \\w is "any word character"; \\W is "anything that is not a word character",
- \\s is "space"; \\S is "not space"

## Replace a string based on a pattern

We can replace a string using `str_replace_all()`. Note that `str_replace()` is also a function, but it will only replace the first instance, and 99.9% of the time it isn't what we want.

```{r}
#will replace every e with a z.
str_replace_all(schs, "e", "z")
#note that this is case sensitive
```

Another useful regex feature is that we can create "groups" with `()`. And we can then access these groups by number

```{r}
test_string <- "two words"

str_replace_all(test_string, "(two) (words)", "\\1")
#\\1 means 'return the stuff in the first set of parens'
```

Let's combine somme of this to select the first word in each school's name:

```{r}
str_replace_all(schs, "^(\\S+).*$", "\\1")
```

This looks a little daunting, but we're:

- Getting the start of the word with `^`
- Getting everything that isn't a space with `\\S`
- Wrapping `\\S` in parentheses to say "this is our first group",
- Selecting "everything not in the first group until the end" with `.*$`
- Returning the first group with `\\1`